Supplemental Materials

### 2024-06-29 products/manuscript/Manuscript_Final.qmd
Merged GT tables; updated supplement
Revisions

## Data: Description and Sources
MAYBE ADD IN HERE ON COVARIATES AND TAKE OUT SOME OTHER PARTS.
In this paper, I analyze a unique sample of 2845 recorded clinical visits to Maternidade Laranjeiras between June 1922 and May 1926. [For information on the data processing, exploratory analysis, and final analysis, including all code in R and the open-access dataset, see my Github repository at BLANK. The `Supplemental_Materials` includes information on data processing, exploratory analysis, and model fitting. All materials, including the dataset, are open access and reproducible, with instructions on the schematic of the workflow.] I extracted the sample from Brazil’s major obstetrics and gynecology journal in the first half of the twentieth century, the *Revista de Gynecologia e d’Obstetricia* (RGO). The RGO was associated with the country’s medical association, the National Academy of Medicine (Academia Nacional de Medicina, ANM) and the Brazilian Society of Obstetrics and Gynecology (Sociedade de Obstetricia e Gynecologia do Brasil), both based in Rio de Janeiro. The journal started in August 1907 as the *Revista de Gynecologia e D’Obstetricia do Rio de Janeiro*. In 1919, it changed to the *Revista de Gynecologia, D’Obstetricia e de Pediatria*. In 1922, it became the *Revista de Gynecologia e D’Obstetricia*.

The journal published obstetricians’ and gynecologists’ clinical observations, analyses of new surgical techniques, and ANM proceedings. Between June 1922 and May 1926, RGO also published the monthly clinical reports of all women treated at the Maternidade Laranjeiras. From the available issues, I recorded the following information, when available, for all patients: patient number, parity, skin color, age, nationality, type of delivery ("natural" or vaginal, "interventionist" or requiring some intervention most often forceps, and "operative," or a cesarean section), maternal outcome (death, discharge, transferal to separate hospital), birth outcome (spontaneous abortion, stillbirth or neonatal death, or live birth), and the mother’s reproductive history [I was unable to locate vol. 18, nos. 4, 5, 6, 8 (1924) and vol. 20, no. 4 (1926). For full citations on the journals from which these data came, see @roth2020; @roth2020a].

@@

I ask: Did the legacy of slavery affect the health of infants born in the first public maternity hospital in Brazil? I quantify how maternal race, nationality, age, and gestational status explain racial disparities in infant birth weight in Rio de Janeiro, Brazil’s first public maternity hospital, the Maternidade Laranjeiras, in the 1920s. I hypothesize that infants born to women of color, defined as mixed-race (*parda*) or Black (*preta*), had lower birth weights than infants born to White women, whether Brazilian or immigrant, given the historical legacies of slavery on population health in Brazil.

# Methods

I use linear models to estimate the relationship of maternal variables on infant birth weight.

## Study Measures

The outcome I will measure is infant birth weight. The World Health Organization (WHO) currently classifies birth weight into the following categories: extremely low (\<999g); very low (1000-1499g); low (1500-2499g); normal (2500-3999g); and high (≥4000g) [@worldhealthorganization2022]. I employ birthweight as a continuous variable, so these categories are not relevant for the linear models but can help understand the skew of the data. For my main models, I measured birthweight as a continuous variable in line with current studies [CITE]. Due to the lack of comprehensive information on gestational age in these data, I do not consider birth weight in reference to gestational age, as is often done in current studies [@cutland2017]. I discuss this constraint in the limitations section below. Infant length was recorded in centimeters.

@@

Spontaneous abortion data points are included in the exploratory analysis but excluded from the statistical analysis. I also excluded any live births weighing \<500 grams to exclude any possible stillbirths not recorded as such [@nyarko2013; @wehby2015; @wehby2017].

@@

@@

I ran four linear models to understand the relationship between maternal factors and infant birth weight. The first two models are a simple and multiple linear regression, respectively, with outcome variable as birthweight in grams. In these two models, I recoded maternal skin color as maternal ancestry (`Euro-descent` or `Afro-descent`).

@@

$$\text{BW} = \alpha + \beta_1 \text{MA} + \epsilon$$ whereas `BW` is birthweight in grams, `MA` is maternal ancestry (Euro-descent, Afro-descent), and $\epsilon$ is the error term. Results are displayed in Table 2.

```{r}
#| label: tbl-table2
#| tbl-cap: "Model 1 Results"
#| echo: FALSE
table2 <- readRDS(here::here("results", "tables", "table2_final.rds"))
flex_table2 <- as_flex_table(table2)
flex_table2 <- flex_table2 %>%
  flextable::set_table_properties(width = .5, align = "center")
flex_table2
```

@@

$$\text{BW} = \alpha + \beta_1 \text{MA} + \beta_2 \text{Gest} + \beta_3 \text{Nat} + \beta_4 \text{Age} + \epsilon$$ whereas `BW` is birthweight in grams, `MA` is maternal ancestry (Euro-descent, Afro-descent), `Gest` is gestational status (nullipara, primipara, multipara, multigravida), `Age` is maternal age in years, and $\epsilon$ is the error term. Results are displayed in Table 3.

```{r}
#| label: tbl-table3
#| tbl-cap: "Model 2 Results"
#| echo: FALSE
i = 16.5
w = i*0.3937
font_size <- 10
table3 <- readRDS(here::here("results", "tables", "table3_final.rds"))
flex_table3 <- as_flex_table(table3)
flex_table3 <- autofit(flex_table3)
auto_widths <- dim(flex_table3)$widths/sum(dim(flex_table3)$widths) # calculate column widths
flex_table3 <- width(flex_table3, width=w*auto_widths) # adjust column widths
flex_table3 <- fontsize(flex_table3, size=font_size) # set font size
flex_table3
```

The second two models are also simple and multiple linear regressions with outcome variable as birthweight in grams. However, in these two models, I maintained maternal skin color as three categories, White (the reference group), Black, and mixed race to see if there is differential outcomes for Black and mixed-race women. The first simple linear regression model is as follows:

$$\text{BW} = \alpha + \beta_1 \text{MC} + \epsilon$$ whereas `BW` is birthweight in grams, `MC` is maternal skin color (White, Black, and mixed race), and $\epsilon$ is the error term. Results are displayed in Table 4.

```{r}
#| label: tbl-table4
#| tbl-cap: "Model 3 Results"
#| echo: FALSE
table4 <- readRDS(here::here("results", "tables", "table4_final.rds"))
flex_table4 <- as_flex_table(table4)
flex_table4 <- flex_table4 %>%
  flextable::set_table_properties(width = .5, align = "center")
flex_table4
```

@@

```{r}
#| label: tbl-table5
#| tbl-cap: "Model 4 Results"
#| echo: FALSE
i = 16.5
w = i*0.3937
font_size <- 10
table5 <- readRDS(here::here("results", "tables", "table5_final.rds"))
flex_table5 <- as_flex_table(table5)
flex_table5 <- autofit(flex_table5)
auto_widths <- dim(flex_table5)$widths/sum(dim(flex_table5)$widths) # calculate column widths
flex_table5 <- width(flex_table5, width=w*auto_widths) # adjust column widths
flex_table5 <- fontsize(flex_table5, size=font_size) # set font size
flex_table5
```

@@

All models based on historical data must contend with limited sample sizes, missing data, and imprecise measures of health and nutrition [@franken2019; @steckel1998; @roberts2014]. In this study, a major limitation was the original published clinical notes' exclusion of gestational age. This means that the birthweight variable is capturing both smaller babies born at term (low weight at-term infants, or small for gestational age) and pre-term infants, or those born before 37 weeks gestation. This is a major limitation, as pre-term infants are more likely to have lower birth weights and are at higher risk for infant mortality [@cutland2017]. Given the absence of accurate technological techniques for determining gestational age in the past, this limitation is inherent to any historical inquiry into birthweights, although other historical studies have datasets that include gestational age [@costa1998; @costa2004; @goldin1989].

>>>>

## Data: Description and Sources

In this paper, I analyze a unique sample of 2845 recorded clinical visits to Maternidade Laranjeiras between June 1922 and May 1926. I extracted the sample from Brazil’s first major obstetrics and gynecology journal, the *Revista de Gynecologia e d’Obstetricia* (RGO). The RGO was associated with the country’s national medical association and obstetrics and gynecological society, both based in Rio de Janeiro. The journal published obstetricians’ and gynecologists’ clinical observations, analyses of new surgical techniques, and professional proceedings. Between June 1922 and May 1926, RGO also published the monthly clinical reports of all women treated at Laranjeiras. [@roth2020; @roth2020a. For information on the data processing, exploratory analysis, and final analysis, including all code in R and the open-access dataset, see my Github repository. The `Supplemental_Materials` includes information on data processing, exploratory analysis, and model fitting. All materials, including the data set, are open access and reproducible, with instructions on the schematic of the workflow available in the README. @roth2024.]

@@

I ask: Did the legacy of slavery affect the health of infants born in the first public maternity hospital in Brazil measured using birth weight as a proxy for general fetal and neonatal health? I quantify how maternal race, nationality, age, and gestational status explain racial disparities in infant birth weight in Rio de Janeiro’s first public maternity hospital. I hypothesize that infants born to women of color, defined as mixed-race (*parda*) or Black (*preta*), had lower birth weights than infants born to White women, whether Brazilian or immigrant, given the historical legacies of slavery on population health in Brazil.

To explore this relationship, I use linear models to estimate the relationship of maternal variables on infant birth weight. In the supplemental analysis, I include results from logistic models that estimated the relationship of maternal variables on infant birth weight as a categorical variable. Most results were insignificant, and thus I exclude them here. The World Health Organization (WHO) currently classifies birth weight into the following categories: extremely low (\<999g); very low (1000-1499g); low (1500-2499g); normal (2500-3999g); and high (≥4000g) [@worldhealthorganization2022]. In my main models, employ birthweight as a continuous variable in line with both current and historical studies [CITE], so these categories are not relevant for the linear models, although they can help explain the skew of the data. Due to the lack of comprehensive information on gestational age in these data, I do not consider birth weight in reference to gestational age, as is often done in current studies [@cutland2017]. I discuss this constraint in the limitations section below. Infant length was recorded in centimeters.

@@

In the exploratory analysis, I included recorded spontaneous abortions and all recorded live births. For the statistical analysis I excluded both spontaneous abortions and any recorded live births weighing \<500 grams to avoid any possible stillbirths [@nyarko2013; @wehby2015; @wehby2017].

@@

ADD IN SECOND SUMMARY TABLE HERE

@@

I ran four linear models to understand the relationship between maternal factors and infant birth weight. The first two models are a simple and multiple linear regression, respectively, with outcome variable as birth weight in grams. In these two models, I recoded maternal skin color as maternal ancestry (`Euro-descent` or `Afro-descent`).

@@

$$\text{BW} = \alpha + \beta_1 \text{MA} + \epsilon$$ whereas `BW` is birth weight in grams, `MA` is maternal ancestry (Euro-descent, Afro-descent), and $\epsilon$ is the error term.

@@

$$\text{BW} = \alpha + \beta_1 \text{MA} + \beta_2 \text{Gest} + \beta_3 \text{Nat} + \beta_4 \text{Age} + \epsilon$$ whereas `BW` is birth weight in grams, `MA` is maternal ancestry (Euro-descent, Afro-descent), `Gest` is gestational status (nullipara, primipara, multipara, multigravida), `Age` is maternal age in years, and $\epsilon$ is the error term. Results for both models are displayed in Table 3.

```{r}
#| label: tbl-table3
#| tbl-cap: "Model 1 and 2 Results"
#| echo: FALSE
table3 <- readRDS(here::here("results", "tables", "table3_final.rds"))
flex_table3 <- as_flex_table(table3)
flex_table3 <- flex_table3 %>%
  flextable::set_table_properties(width = .5, align = "center")
flex_table3
```

The second two models are also simple and multiple linear regressions with outcome variable as birth weight in grams. However, in these two models, I maintained maternal skin color as the original three categories recorded in the hospital records, White (the reference group), Black, and mixed race to see if there is differential outcomes for Black and mixed-race women. The first simple linear regression model is as follows:

$$\text{BW} = \alpha + \beta_1 \text{MC} + \epsilon$$ whereas `BW` is birth weight in grams, `MC` is maternal skin color (White, Black, and mixed race), and $\epsilon$ is the error term. Results are displayed in Table 4.

@@

```{r}
#| label: tbl-table4
#| tbl-cap: "Model 3 and 4 Results"
#| echo: FALSE
table4 <- readRDS(here::here("results", "tables", "table4_final.rds"))
flex_table4 <- as_flex_table(table4)
flex_table4 <- flex_table4 %>%
  flextable::set_table_properties(width = .5, align = "center")
flex_table4
```

@@

All models based on historical data must contend with limited sample sizes, missing data, and imprecise measures of health and nutrition [@franken2019; @steckel1998; @roberts2014]. In this study, a major limitation was the original published clinical notes' exclusion of gestational age. Thus, the birth weight variable captured both smaller babies born at term (low weight at-term infants, or small for gestational age) and pre-term infants, or those born before 37 weeks gestation. This is a major limitation, as pre-term infants are more likely to have lower birth weights and are at higher risk for infant mortality [@cutland2017]. Given the absence of accurate technological techniques for determining gestational age in the past, this limitation is inherent to any historical inquiry into birthweights, although other historical studies have datasets that include gestational age [@costa1998; @costa2004; @goldin1989].

### 2024-06-29 products/manuscript/supplement/Supplemental_Materials.qmd
Merged GT tables; updated supplement
Revisions

@@

```{r}
#| label: tbl-table2
#| tbl-cap: "Supplemental Table"
#| echo: FALSE
table2 <- readRDS(here::here("results", "tables", "table2_final.rds"))
table_gt <- gtsummary::as_gt(table2)
table_gt |> gt::as_raw_html()
```

@@

```{r}
#| label: tbl-table_age
#| tbl-cap: "Supplemental Table"
#| echo: FALSE
table_age <- readRDS(here::here("results", "tables", "tableage_final.rds"))
table_gt <- gtsummary::as_gt(table_age)
table_gt |> gt::as_raw_html()
```

The third and final exploratory statistical model is a multilinear regression, looking at the relationship of maternal skin color, age, nationality, and gestational status on infant birth weight.

```{r}
#| label: tbl-table4
#| tbl-cap: "Supplemental Table"
#| echo: FALSE
table4 <- readRDS(here::here("results", "tables", "table4.rds"))
table_gt <- gtsummary::as_gt(table4)
table_gt |> gt::as_raw_html()
```

In the first and third models, there appears to be an association between maternal skin color and infant birth weight, with Euro-descended women (the reference group) having infants with higher birth weights than Afro-descended women. In the second model, older mothers are associated with giving birth to infants with higher birth weights.

@@

@@

Here I include the cross-validation results for each model.

@@

>>>>

## Models

### Linear Models

@@

```{r}
#| label: tbl-table1
#| tbl-cap: "Model 1"
#| echo: FALSE
table1 <- readRDS(here::here("results", "tables", "table_fit1.rds"))
table_gt <- gtsummary::as_gt(table_fit1)
table_gt |> gt::as_raw_html()
```

@@

```{r}
#| label: tbl-table2
#| tbl-cap: "Model 2"
#| echo: FALSE
table2 <- readRDS(here::here("results", "tables", "table_fit2.rds"))
table_gt <- gtsummary::as_gt(table_fit2)
table_gt |> gt::as_raw_html()
```

The third exploratory statistical model is a multilinear regression, looking at the relationship of maternal skin color, age, nationality, and gestational status on infant birth weight.

```{r}
#| label: tbl-table3
#| tbl-cap: "Model 3"
#| echo: FALSE
table3 <- readRDS(here::here("results", "tables", "table_fit3.rds"))
table_gt <- gtsummary::as_gt(table_fit3)
table_gt |> gt::as_raw_html()
```
In the first and third models, there appears to be an association between maternal skin color and infant birth weight, with Euro-descended women (the reference group) having infants with higher birth weights than Afro-descended women. In the second model, older mothers are associated with giving birth to infants with higher birth weights.

@@

The fourth exploratory model again runs a multivariate linear regression, this time using skin color as it was differentiated in the original data, or with three categories (White, Black, and Mixed Race).

```{r}
#| label: tbl-table4
#| tbl-cap: "Model 4"
#| echo: FALSE
table4 <- readRDS(here::here("results", "tables", "table_fit4.rds"))
table_gt <- gtsummary::as_gt(table_fit4)
table_gt |> gt::as_raw_html()
```

### Logistic Models

Now we will still consider `Birthweight` as the outcome of interest, but we will transform it into a categorical outcome of <=2500 (low birth weight, LBW) and >2500 (normal birth weight, NBW) and fit a logistic model, using the main predictor of interest, `ModifiedColor`.

The first model is a simple logistic regression, with modified maternal skin color as the predictor variable.

```{r}
#| label: tbl-table5
#| tbl-cap: "Model 5"
#| echo: FALSE
table5 <- readRDS(here::here("results", "tables", "table_log1.rds"))
table_gt <- gtsummary::as_gt(table_log1)
table_gt |> gt::as_raw_html()
```

The second model is also a simple logistic regression, this time using `Color` as the predictor variable (so three categories, Black, mixed race, and White).

```{r}
#| label: tbl-table6
#| tbl-cap: "Model 6"
#| echo: FALSE
table6 <- readRDS(here::here("results", "tables", "table_log2.rds"))
table_gt <- gtsummary::as_gt(table_log2)
table_gt |> gt::as_raw_html()
```

Then, I ran two more logistic models, one with `ModifiedColor` and the other covariates `Age`, `ModifiedStatus`, `ModifiedNationality`, and one with `Color` and these same covariates. The results are below.

```{r}
#| label: tbl-table7
#| tbl-cap: "Model 7"
#| echo: FALSE
table7 <- readRDS(here::here("results", "tables", "table_log3.rds"))
table_gt <- gtsummary::as_gt(table_log3)
table_gt |> gt::as_raw_html()
```

```{r}
#| label: tbl-table8
#| tbl-cap: "Model 8"
#| echo: FALSE
table8 <- readRDS(here::here("results", "tables", "table_log4.rds"))
table_gt <- gtsummary::as_gt(table_log4)
table_gt |> gt::as_raw_html()
```
In these two models, the only significant finding is the final logistic model, in which infants born to mixed-race mothers had 21% higher odds of being born with low birth weight compared to infants born to white mothers. The other models did not show significant results.

Best practice is to not categorize continuous variables, as this can include arbitrary cutoffs for variables that are best understood continuously [@Senn2005]. Thus, I will not include these logistic regressions in the final analysis.

@@

Here I include the cross-validation results for the linear models.

@@

The performance metrics for the logistic models are available in the `introanalysis-v1.qmd` file.
